A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2014-04-24 jacobian_calc_CMAME.tex
Numerous changes

\subsubsection{Implementing the complex-step method in \emph{Peridigm}} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Peridigm} in a manner that follows closely the implementation of the FD technique.  However, because of \emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the \emph{Epetra} package, special steps had to be followed to allow the use of complex number data types.  This is because \emph{Epetra} vectors, which can be thought of as distributed memory parallel cousins to standard C++ Standard Template Library (STL) vectors, are hard coded to be of type {\tt double} and can not be simply declared as having a complex data type. \todo[]{Re: I.3} For the reader's convenience, \emph{Epetra} vectors and STL vectors are what are known as 'container classes', whose purpose in object oriented computer programming is to automate the creation, accessing, modification and destruction of the underlying contained data. Templated container classes allow for user choice in what properties that underlying data exhibits.  For example, in this work, complex arithmetic cannot be performed on the data contained within an \emph{Epetra} vector because that data, of type {\tt double} does not have the property of following the rules of complex arithmetic. The reason the non-templated \emph{Epetra} vector could not be dispensed with for the purposes of this work is because the \emph{Epetra} vector was tightly integrated with \emph{Peridigm's} parallel communication strategy and linear algebra code. Program methods that were used in the course of applying the FD method were copied, renamed and re-written to follow the CS formula.  Because of restrictions imposed by the container class being used, this involved changing the data type of intermediate variables which were locally scoped, and dynamically casting persistent memory variables to complex data types where necessary. \todo[]{Re: I.3}For the reader's convenience, 'locally scoped' in the C++ programming language refers to data which is created within a context more specific than the context of the entire program, or 'translation unit'.  Most data created this way is destroyed when that specific context is left. The topic of contexts in C++ is very broad, so for the purposes of this work context refers to a program method or function that begins, concludes and yields a result, also destroying locally scoped data. Dynamic casting for the purposes of this work refers to copying the value of data contained within an \emph{Epetra} vector into locally scoped data that has the property of following the rules of complex arithmetic, where a mapping called 'type casting' allows the translation of data between different types. No attempt was made to ``tune'' the CS code in any way for performance. The algorithm  was implemented in such as way that it was in one-to-one correspondence with the FD and CD methods; however, it must be noted that the overhead associated with the dynamic casts in the CS method likely hurts its overall performance and this issue could be alleviated by using the next generation of templated \emph{TPetra} vectors available in \emph{Trilinos} that are capable of taking an explicit instantiation of any data type including {\tt complex}. \todo[]{Re: I.3}For the purposes of this work, an explicit instantiation refers to requesting the creation of an object, or instance, of a templated class or container class that holds data which has the property of following desired mathematic or logic rules. The version of \emph{Peridigm} modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computer scientists; therefore,
many advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic, and distributed data structures, \todo[]{Re: I.3}for example \emph{Epetra} vectors, were used in the basic
    code and by necessity in the modified code. These concepts are not vital to
understanding the complex step method itself, but may be necessary to understand complex
step in \emph{Peridigm}. To see CS and other methods in a more narrow context,
it is advised that readers also take a look at the simple examples provided on
the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each
method would solve a problem concurrently. In this context, concurrently means
successively, drawing from the same independent variables, yet within the same
computational process; this definition is characterized in further detail in
Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by
$\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing
tension along the axis parallel to the long dimension of the block. The test
problems were set up in \emph{Peridigm} as quasi-static equilibrium problems,
where the displacement boundary conditions used to apply tension were applied
gradually in ``load steps'', and an equilibrium solution for each load step was
achieved before applying the next load step. Applying the load
in gradual steps is a widely used technique to ensure the Newton's method is
always initialized near the actual solution and therefore likely to converge.
The fictional material represented in the model had values of
$\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$
for shear modulus. The peridynamic horizon, a length scale that effectively
sets the bandwidth of tangent-stiffness matrix in the computation, used in the test
problems was always set to three times the nominal discretization size (i.e.
node spacing in the particle discretization scheme). This results in a minimum
tangent-stiffness matrix bandwidth of 21 for a 3D problem (for an optimal node
numbering scheme). Pictured in Fig.~\ref{fig:SCMesh} is a viewgraph for one of
the peridynamic meshes used in this study. The coloration in the image
indicates node numbering, where a smooth color transition means that the nodes
have close ID numbers. The abrupt color changes in the center and at the ends
of the body show that numbering is not contiguous and is assigned with respect to four regions. The purpose of designating
these regions, for this study, was to apply displacement boundary
conditions called \emph{volume constraints} in peridynamics.

\begin{figure}[tbp] 
    \centering 
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh2000_replacement.png}
        \caption{2000 node mesh.}
        \label{fig:MCMesh}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh1000000_replacement.png}
        \caption{1000000 node mesh.}
        \label{fig:SCMesh}
    \end{subfigure}
    \caption{Peridynamic discritizations used in study.}
    \label{fig:pd_discrete}
\end{figure}

A series of nine single compute core (i.e.\ serial) test problems were run with
the discretization size being refined with every test in the series. The aim of
increasing the refinement was to examine differences between the methods in
terms of accuracy and speed. This series would comprise the single core runs. 
\todo[]{Re: II.16}
The specific parameters of these tests can be found in \emph{Peridigm} input
{\tt xml} files included in the paper repository, however the parameters
are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow
users with the appropriately modified version of \emph{Peridigm} to reproduce
the results found in this paper.

\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Test Parameters} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon & Load rate & Number Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\meter}& \si{\centi\meter\per\second} &   &     \\
        \midrule
        1000 & $1.99$             & $.3$   & $2.0$ & $2$ & $\num{10E-8}$ \\
        2000 & $4.25$             & $.24$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        3000 & $7.79$             & $.21$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        4000 & $15.4$             & $.19$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        6000 & $20.0$             & $.16$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        8000 & $31.4$             & $.15$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        12000 & $40.1$            & $.131$ & $2.0$ & $2$ & $\num{10E-8}$ \\
        16000 & $83.9$            & $.12$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        32000 & $165.0$           & $.095$ & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $.03$  & $2.0$ & $2$ & $\num{10E-8}$ \\
        \bottomrule
    \end{tabular}
\end{table*}

Another series of four test problems were simulated, however this time the
number of compute cores used to solve the problem was increased from test to
test while the discretization level was held constant at 1 million
computational nodes (3 million degrees of freedom). This allowed for sufficient
computational nodes per core even at the highest level of parallelization such
that message passing computation did not overwhelm the simulation.  The
specific parameters of these tests can be found in \emph{Peridigm} input {\tt
xml} files also included in the paper repository, but a summary of the most relevant
ones appears in Table~\ref{tab:TestParams}. A viewgraph of the 1 million node mesh is
shown in Fig.~\ref{fig:MCMesh}. The nodes in the image, although actually discrete points
in space are dense enough to appear as a continuous body at a distance.

\todo[]{Re: I.3}For the purposes of this work message passing is what
synchronizes multiple computers that each are working on a separate section
of the problem domain. This operation involves the time-consuming use of
the data network. To reduce the proportion of simulation time spent on
message passing, good practice is to maximize the amount of information
contained within a subdomain solely managed by a single computer, relative
to the amount of information shared across multiple subdomains and
therefor multiple computers. A useful experiment to illustrate this idea
concretely would be to implement a code to solve a 2D Laplace equation for
simulating a simple heat conduction problem on two processors. Use the
textbook finite difference method, which in this case does not refer just
to the differentiation technique.  First decompose the domain into two
contiguous chunks, having one shared border, and distribute these chunks,
or subdomains to the two processors. Perform the simulation a multitude of
times to calculate an average runtime.  Next implement the solution a
different way, instead partitioning the problem domain into many narrow,
alternating strips, and distribute those to again two processors, where adjacent
strips in the graph of the domain should be sent to different processors. This way
of decomposing the domain will cause much overlap between the subdomains
managed by the two processors and much time will be spent synchronizing the
subdomain overlap or 'ghost nodes' between the processors. Run the modified
simulation a multitude of times and take the average runtime. The larger
the unpartitioned domain, the worse the discrepancy should be.

\subsubsection{Quantities of interest} \label{JGAM}

As stated in the Introduction, a goal of this study was to compare CS, FD, and
CD on the basis of accuracy. AD was omitted from the comparison because it
served as the standard of accuracy for the other methods in the absence of
appropriate analytical forms for the tangent-stiffness matrix associated with
the system solved in this study. The assumption that AD is accurate enough to
serve as a standard is supported by AD's implementation as a computerized chain
rule as explained in Section~\ref{ADsubsection}.

@@

The other goal was to compare the four methods, including AD, on the basis of
speed. Speed is defined as the total compute-time of one iteration of a Newton
step which includes the tangent-stiffness matrix calculation and parallel assembly.
The speed of iteration was measured because it could be so done at the same
time as accuracy was being measured given a single tangent-stiffness
calculation.  It was assumed that the order that each method was evaluated in
parallel was unimportant, that evaluating each method successively within each
solver iteration did not affect their performance individually and that speed
of computation did not change with time. These assumptions allowed the test
program to run the same problem with each of the methods at effectively the
same time and generate an equal volume of data from each method. It was also
assumed that the tangent-stiffness matrix calculation time for each of the
methods did not vary based on the current configuration, as it changes slightly
from iteration to iteration as the Newton solution iterate is updated.  This
assumption allowed calculation time measurements to be averaged over all
iterations within a single load step. The purpose of averaging calculation time
measurements was to informally address the extraneous variables of parallel
evaluation order and computer system load due to system processes not
associated with this study. While not part of this study, the example programs
available in the paper repository allow the reader to make a comparison of the
methods for themselves on the basis of accuracy and convergence rate for two
example nonlinear systems. In these examples, one will notice that the finite
difference methods will lose quadratic convergence when the step size selected
is too large such that accuracy relative to AD is decreased.

The goals of collecting accuracy and speed measurements were achieved by
developing and implementing metrics within the simulation program used in the
study. The metric used to measure the accuracy of a tangent-stiffness matrix
was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix)
of the element-wise difference between the tangent-stiffness matrix produced by
the method being evaluated and the tangent-stiffness matrix produced by the AD
based method (i.e., the \emph{exact} derivatives), given that both methods were
set upon the same problem. The lower the value of this metric, the more
accurate the method. The expression for the accuracy metric was
%
\begin{equation} D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2},
\label{eqn:accuracy} \end{equation}
%
where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the
AD based method, and $M$ is replaced with either CS, FD, or CD depending on the
method being compared.  The metric used to compare the speed of the different
methods was the time in seconds required to calculate the tangent-stiffness
matrix. A final basis of comparison used in this study called
\emph{computational efficiency} is based on specific calculation time per
tangent-stiffness matrix element.

\subsubsection{Computers and other software} 
%
Discretization for the test problems run in this study was done using the
software package \emph{Cubit} developed at Sandia National Laboratories
\cite{ref-Cubit}. \emph{Cubit} generates a finite element mesh that is
internally converted to a ``particle'' discretization internally when calling
\emph{Peridigm}. The number of computational nodes in \emph{Peridigm}
corresponds to the number of finite elements in the discretization, not the
number of finite element nodes.  An example journal script can be found on the
corresponding author's website.

The test problems were run on the \emph{Stampede} HPC cluster computer housed
at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at
Austin. The single core test series was run using the {\tt normal} queue
compute nodes, while the parallel test series was run using one to four {\tt
large memory} nodes each having one terabyte of local RAM. This large amount of
memory was required for reasons explained in Section~\ref{JGAM}, second paragraph.

\subsubsection{Data reduction}

The output produced by \emph{Peridigm} including the additional accuracy and
speed measurements was redirected from the console to text files by the
resource manager, \emph{SLURM} (Simple Linux User Resource Manager).
Directories and filenames were chosen so that data would be easy to sort by
individual test run.  \emph{Python} scripts were then used to post-process and
plot the data. These scripts averaged accuracy or speed data for a tests
corresponding to a particular peridynamic node density over all iterations for
that run, and then plotted this averaged data  as a function of peridynamic
node density as it varied from test to test in the series.  Similarly, for the
parallel tests, the data was averaged and plotted in much the same way, but as
a function of number of compute cores used to solve the problem. Confidence
intervals were not calculated on the measurements. The main reason is that
network traffic from other users makes running tests on the HPC cluster a time
varying process. These data reduction and plotting scripts can be found on the
corresponding author's website.

\section{Results and discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by Eqn.~(\ref{eqn:accuracy}) and the choice of elastic modulii mentioned in Section~\ref{tcs}.
%
\begin{table*}[!ht]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

\subsection{Speed data}

>>>>

\subsubsection{Implementing the complex-step method in Peridigm} 
\label{subsubsec:Impl}
%
The CS method was implemented in \emph{Peridigm} in a manner that follows closely the implementation of the FD technique.  However, because of \emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the \emph{Epetra} package, special steps had to be followed to allow the use of complex number data types.  This is because \emph{Epetra} vectors, which can be thought of as distributed memory parallel cousins to standard C++ Standard Template Library (STL) vectors, are hard coded to be of type {\tt double} and can not be simply declared as having a complex data type. \todo[]{Re: I.3} \emph{Epetra} vectors and STL vectors are what are known as \emph{container classes}, whose purpose in object-oriented computer programming is to automate the creation, accessing, modification,  and destruction of the underlying contained data.  \emph{Templated} container classes allow for user choice in what properties that underlying data exhibits.  For example, in this work, complex arithmetic cannot be performed on the data contained within an \emph{Epetra} vector because that data, of type {\tt double},  does not have the property of following the rules of complex arithmetic. The reason the non-templated \emph{Epetra} vector could not be dispensed with for the purposes of this work is because the \emph{Epetra} vectors are tightly integrated with \emph{Peridigm}'s parallel communication strategy and linear algebra code. Because of restrictions imposed by the container class being used, this involved changing the data type of intermediate variables which were locally scoped (i.e. only allocated to memory within the specific CS method and later destroyed), and dynamically casting persistent memory variables to complex data types where necessary. No attempt was made to ``tune'' the CS code in any way for performance. The algorithm  was implemented in such as way that it was in one-to-one correspondence with the FD and CD methods; however, it must be noted that the overhead associated with the dynamic casts in the CS method likely hurts its overall performance and this issue could be alleviated by using the next generation of templated \emph{TPetra} vectors available in \emph{Trilinos} that are capable of taking an explicit instantiation of any data type including {\tt complex}. The version of \emph{Peridigm} modified to implement CS can be found in the paper repository.

\emph{Peridigm} was primarily written by computational scientists and engineers; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures are employed. These concepts are not vital to understanding the CS method itself, but may be necessary to understand CS in \emph{Peridigm}. To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples, such as that described in Section~\ref{sec:WorkingExamples},  provided on the corresponding author's website.

@@

The methods were compared by running two sets of test problems where each method would solve a problem concurrently. In this context, concurrently means successively, drawing from the same independent variables, yet within the same computational process; this definition is characterized in further detail in Section~\ref{JGAM}. Each test problem simulated a $\SI{4}{\meter}$ by $\SI{0.5}{\meter}$ by $\SI{0.5}{\meter}$ meter block of a material undergoing tension along the axis parallel to the long dimension of the block. The test problems were set up in \emph{Peridigm} as quasi-static equilibrium problems, where the displacement boundary conditions used to apply tension were applied gradually in ``load steps'', and an equilibrium solution for each load step was achieved before applying the next load step. Applying the load in gradual steps is a widely used technique to ensure the Newton's method is always initialized near the actual solution and therefore likely to converge.  The fictional material represented in the model had values of $\SI{1.515e4}{\mega\pascal}$ for bulk modulus, and $\SI{7.813e4}{\mega\pascal}$ for shear modulus. The peridynamic horizon, a length scale that effectively sets the bandwidth of tangent-stiffness matrix in the computation, used in the test problems was always set to three times the nominal discretization size (i.e.  node spacing in the particle discretization scheme). Pictured in Figure.~\ref{fig:pd_discrete} is an illustration of the peridynamic discretizations used in this study. The coloration in the images indicate material regions that were defined to assist in post-processing application of boundary conditions. 
%
\begin{figure}[tbp] 
    \centering 
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh2000_replacement.png}
        \caption{2000 node mesh.}
        \label{fig:MCMesh}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=\textwidth]{./figs/mesh1000000_replacement.png}
        \caption{1000000 node mesh.}
        \label{fig:SCMesh}
    \end{subfigure}
    \caption{Peridynamic discritizations used in study.}
    \label{fig:pd_discrete}
\end{figure}

A series of nine single compute core (i.e.\ serial) test problems were run with the discretization size being refined with every test in the series. The aim of increasing the refinement was to examine differences between the methods in terms of accuracy and speed. This series would comprise the single core runs.  \todo[]{Re: II.16} The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files included in the paper repository, however the parameters are reproduced in Table~\ref{tab:TestParams}. These {\tt xml} files allow users with the appropriately modified version of \emph{Peridigm} to reproduce the results found in this paper.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
        \caption{Test Parameters} \label{tab:TestParams}   
       \begin{tabular}{c c c c c c}
         \toprule
         Peri. nodes & Nozero TS elems. & Horizon ($\delta$) & Load rate & Number Load steps & Abs. tolerance \\ 
                     & $\times 10^6$    & \si{\centi\meter}& \si{\centi\meter\per\second} &   &     \\
        \midrule
        1000 & $1.99$             & $30$   & $2$ & $2$ & $\num{10E-8}$ \\
        2000 & $4.25$             & $24$  & $2$ & $2$ & $\num{10E-8}$ \\
        3000 & $7.79$             & $21$  & $2$ & $2$ & $\num{10E-8}$ \\
        4000 & $15.4$             & $19$  & $2$ & $2$ & $\num{10E-8}$ \\
        6000 & $20.0$             & $16$  & $2$ & $2$ & $\num{10E-8}$ \\
        8000 & $31.4$             & $15$  & $2$ & $2$ & $\num{10E-8}$ \\
        12000 & $40.1$            & $13.1$ & $2$ & $2$ & $\num{10E-8}$ \\
        16000 & $83.9$            & $12$  & $2$ & $2$ & $\num{10E-8}$ \\
        32000 & $165.0$           & $9.5$ & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        1000000 & $1670$          & $3$  & $2$ & $2$ & $\num{10E-8}$ \\
        \bottomrule
    \end{tabular}
\end{table*}

Another series of four test problems were simulated, however this time the number of compute cores used to solve the problem was increased from test to test while the discretization level was held constant at 1 million computational nodes (3 million degrees of freedom). This allowed for sufficient computational nodes per core even at the highest level of parallelization such that message passing computation did not overwhelm the simulation.  The specific parameters of these tests can be found in \emph{Peridigm} input {\tt xml} files also included in the paper repository, but a summary of the most relevant ones appears in Table~\ref{tab:TestParams}. An image of the 1 million node discretization is shown in Figure~\ref{fig:MCMesh}. The nodes in the image, although actually discrete points in space are dense enough to appear as a continuous body at a distance.

\subsubsection{Quantities-of-interest} 
\label{JGAM}

As stated in Section~\ref{sec:intro}, a goal of this study was to compare CS, FD, and CD on the basis of accuracy. AD was omitted from the comparison because it served as the standard of accuracy for the other methods in the absence of appropriate analytical forms for the tangent-stiffness matrix associated with the system solved in this study. The assumption that AD is accurate enough to serve as a standard is supported by AD's implementation as a computerized chain rule as explained in Section~\ref{ADsubsection}.

@@

The other goal was to compare the four methods, including AD, on the basis of speed. Speed is defined as the total compute-time of one iteration of a Newton step which includes the tangent-stiffness matrix calculation and parallel assembly.  The speed of iteration was measured because it could be so done at the same time as accuracy was being measured given a single tangent-stiffness calculation.  It was assumed that the order that each method was evaluated in parallel was unimportant, that evaluating each method successively within each solver iteration did not affect their performance individually and that speed of computation did not change with time. These assumptions allowed the test program to run the same problem with each of the methods at effectively the same time and generate an equal volume of data from each method. It was also assumed that the tangent-stiffness matrix calculation time for each of the methods did not vary based on the current configuration, as it changes slightly from iteration to iteration as the Newton solution iterate is updated.  This assumption allowed calculation time measurements to be averaged over all iterations within a single load step. The purpose of averaging calculation time measurements was to informally address the extraneous variables of parallel evaluation order and computer system load due to system processes not associated with this study. While not part of this study, the example programs available in the paper repository allow the reader to make a comparison of the methods for themselves on the basis of accuracy and convergence rate for two example nonlinear systems. In these examples, one will notice that the finite difference methods will lose quadratic convergence when the step size selected is too large such that accuracy relative to AD is decreased as shown in Table~\ref{tab:ConvergenceStudy}.

The goals of collecting accuracy and speed measurements were achieved by developing and implementing metrics within the simulation program used in the study. The metric used to measure the accuracy of a tangent-stiffness matrix was the Frobenius norm (analogous to an $l^2$-norm except defined on a matrix) of the element-wise difference between the tangent-stiffness matrix produced by the method being evaluated and the tangent-stiffness matrix produced by the AD based method (i.e., the \emph{exact} derivatives to numerical precision), given that both methods were set upon the same problem. The lower the value of this metric, the more accurate the method. The expression for the accuracy metric was
%
\begin{equation} 
    D = \sqrt{\sum_i \sum_j(K^{AD}_{ij} - K^{M}_{ij})^2},
    \label{eqn:accuracy} 
\end{equation}
%
where $D$ is distance, $K^{AD}$ is the tangent-stiffness matrix produced by the AD based method, and $M$ is replaced with either CS, FD, or CD depending on the method being compared.  The metric used to compare the speed of the different methods was the time in seconds required to calculate the tangent-stiffness matrix. A final basis of comparison used in this study called \emph{computational efficiency} is based on specific calculation time per tangent-stiffness matrix element.

\subsubsection{Computers and other software} 
%
Discretization for the test problems run in this study was done using the software package \emph{Cubit} developed at Sandia National Laboratories \cite{ref-Cubit}. \emph{Cubit} generates a finite element mesh that is converted to a ``particle'' discretization internally when running \emph{Peridigm}. The number of computational nodes in \emph{Peridigm} corresponds to the number of finite elements in the discretization, not the number of finite element nodes.  An example journal script can be found in the corresponding paper repository.

The test problems were run on the \emph{Stampede} HPC cluster computer housed at \emph{TACC} (Texas Advanced Computing Center) at The University of Texas at Austin. The single core test series was run using the {\tt normal} queue compute nodes, while the parallel test series was run using one to four {\tt large memory} nodes each having one terabyte of local RAM. This large amount of memory was required for reasons explained in Section~\ref{JGAM}.

\subsubsection{Data reduction} 
The output produced by \emph{Peridigm} including the additional accuracy and speed measurements was redirected from the console to text files by the resource manager, \emph{SLURM} (Simple Linux User Resource Manager).  Directories and filenames were chosen so that data would be easy to sort by individual test run.  \emph{Python} scripts were then used to post-process and plot the data. These scripts averaged accuracy or speed data for a tests corresponding to a particular peridynamic node density over all iterations for that run, and then plotted this averaged data  as a function of peridynamic node density as it varied from test to test in the series.  Similarly, for the parallel tests, the data was averaged and plotted in much the same way, but as a function of number of compute cores used to solve the problem. Confidence intervals were not calculated on the measurements. The main reason is that network traffic from other users makes running tests on the HPC cluster a time varying process. These data reduction and plotting scripts can be found in the paper repository.

\section{Results and discussion}
\label{sec:Res}
%
Table~\ref{tab:results} shows the compute core count, number of nonzero tangent-stiffness (TS) matrix elements, load step average calculation time and load step average accuracy. The accuracy measure is given by equation~(\ref{eqn:accuracy}) and the choice of elastic modulii mentioned in Section~\ref{tcs}.
%
\begin{table*}[!tbp]    
  \scriptsize
  \centering
        \caption{Averaged Results, Each Test} \label{tab:results}   
       \begin{tabular}{c c c c c c c c c}
         \toprule
         Cores & TS Elements & \multicolumn{4}{c}{Calculation Time ($\si{\second}$)} & \multicolumn{3}{c}{Accuracy Difference ($\si{\mega\pascal}$)} \\ 
         & $\times 10^6$ & CS & CD & FD & AD & CS & CD & FD \\
        \midrule
        1 & $\num{1.99}$  & 3.5 & 2.7 & 1.7 & 1.6 & $\num{1.92E-10}$ & $\num{1.21E-4}$ & .137 \\
        1 & $\num{4.25}$  & 6.2& 4.9& 3.2& 2.9 & $\num{2.28E-10}$ & $\num{9.94E-4}$ & .148 \\
        1 & $\num{7.79}$  & 11.2& 8.9& 5.7& 5.2 & $\num{2.38E-10}$ & $\num{1.59E-4}$ & .145\\
        1 & $\num{15.4}$  & 26.7& 21.0& 13.4& 12.2 & $\num{2.33E-10}$ & $\num{4.61E-4}$ & .12 \\
        1 & $\num{20.0}$  & 28.1& 22.2& 14.4& 13.1 & $\num{2.76E-10}$ & $\num{1.05E-3}$ & .145 \\
        1 & $\num{31.4}$  & 47.6& 37.7& 24.2& 21.9 & $\num{2.64E-10}$ & $\num{1.65E-3}$ & .133 \\
        1 & $\num{40.1}$  & 55.6& 44.1& 28.4& 25.9 & $\num{3.03E-10}$ & $\num{1.92E-3}$ & .148 \\
        1 & $\num{83.9}$  & 138.9& 109.6& 70.2& 64.0 & $\num{3.63E-10}$ & $\num{1.64E-3}$ & .123 \\
        1 & $\num{165.0}$  & 277.3& 218.1& 139.4& 126.5 & $\num{3.26E-10}$ & $\num{2.18E-3}$ & .128 \\
        32 & $\num{1670}$  & 336.1& 277.7& 200.6& 233.1 & $\num{6.21E-10}$ & $\num{1.52E-2}$ & .176 \\
        64 & $\num{1670}$  & 169.9& 140.7& 102.0& 119.7 & $\num{6.20E-10}$ & $\num{1.50E-2}$ & .177 \\
        96 & $\num{1670}$  & 114.7& 95.0& 69.1 & 79.7 & $\num{6.18E-10}$ & $\num{1.50E-2}$ & .177 \\
        128 & $\num{1670}$  & 86.4& 71.8& 52.4 &58.8 & $\num{6.16E-10}$ & $\num{1.47E-2}$ & .177 \\
        \bottomrule
    \end{tabular}
\end{table*}

\subsection{Speed data} 
%
Speed measurements were taken according to the methods mentioned in Section~\ref{tcs}. The averaged results of these speed measurements for the serial tests can be seen in Fig.~\ref{fig:serial_speed}.  Average time in seconds to compute a TS matrix is plotted as a function of the number of nonzero TS elements.  CS was the slowest of the four methods, taking the longest average time per iteration, followed by CD, FD and AD in that order. Calculation time has a power-law relationship with problem size, with a power-law index of roughly $\num{1.0E-6}$ for each of the methods. This bodes well for the scalability of each of the methods by themselves in \emph{Peridigm}. If we assumed every individual operation, generally speaking, took the same amount of time on average such that compute time were a measure of number of operations, this suggests that the methods as in \emph{Peridigm} have the same complexity class when considering computing a TS matrix alone, that is computation time is $\mathcal{O}(10^{(10^{-6} \times \log_{10}n)})$. The definition of complexity class used here is the informal one given in \cite[pp. 1059]{cormen2009introduction}. This would suggest that the methods are equally well suited to the task of computing a TS matrix, if it were further assumed that the TS matrices produced by each method were identical. This is not the case as seen in the following subsection.
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
    \caption{Serial test series speed measurements.} 
    \label{fig:serial_speed}
\end{figure}
%
For the parallel simulations, averaged speed results appear in Figure~\ref{fig:multi_speed}. For this series, rather than varying the number of nonzero TS matrix elements, the number of compute cores used to solve the problem is increased, while the number of nonzero TS matrix elements remained constant at $\num{1.67E9}$ .  CS again was the slowest method, followed by CD, AD and then FD.  There is a somewhat surprise reversal of FD and AD for these parallel simulations.  Figure~\ref{fig:multi_speed} indicates that when looking at average iteration time, the factor of speed improvement scales inversely with the number of compute cores used for the given conditions of this group of tests. Figure~\ref{fig:multi_speed} clearly shows that when core count doubles, compute time halves, for each method. Again, these results show that each of the methods as implemented in \emph{Peridigm} scale well when additional processors are used. 
%
\begin{figure}[tbp] 
    \centering 
    \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
    \caption{Multicore test series speed measurements.} 
    \label{fig:multi_speed}
\end{figure}
%
A speculative explanation as to why FD and AD have reversed places in the speed tests with respect to the serial results is that the AD data-structures carry a list of partial derivatives of the functional evaluations along side the functional evaluations themselves only to be processed at the end of the tangent-stiffness matrix calculation when the derivatives are requested. This represents additional load on the hardware memory subsystem, such that may exceed that of FD yet still falls under a supposed memory bandwidth threshold during serial operation. Cores within a multicore processor use the same shared bus to access memory, such that cores compete for memory access. Supposing that exceeding the memory bandwidth threshold penalizes additional requested bandwidth, it is likely that the decrease in available memory bandwidth per core caused by parallel operation coupled with AD's additional resource requirements causes the AD method to be slower than FD.
